### 2016-04-19 thesis_paper/source.tex
paper stuff

\documentclass[10pt,journal,final,letterpaper,twocolumn]{IEEEtran}
%\documentclass{ssdbm}
\usepackage{wrapfig,subfigure,setspace,verbatim,cite,url,graphicx,float}
\usepackage{algorithmic,amsfonts,amssymb,amsmath,calc,calrsfs,cleveref}
\usepackage[justification=centering]{caption}
%\usepackage{fixltx2e}

\restylefloat{table}

% This was from the NEW paper
%----------------------------------------------------------
\RequirePackage{fix-cm}

\DeclareMathOperator{\var}{var}
\newcounter{qcounter}

%\smartqed  % flush right qed marks, e.g. at end of proof
%----------------------------------------------------------

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\newcommand{\SEC}[1] {Section \ref{#1}}
\newcommand{\Eq}[1] {Equation \ref{#1}}
\newcommand{\eq}[1] {Eq. (\ref{#1})}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

% paper title
\title{Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU}

% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs

\author{Iliiazbek~Akhmedov, Yi-Cheng~Tu~$\dagger$, Vladimir Grupcev, Joseph Fogarty, Sagar Pandit % <-this % stops a space

%\thanks{* These authors contributed equally to this work.}
\thanks{$\dagger$ Author to whom all correspondence should be sent.}
\thanks{Iliiazbek Akhmedov, Vladimir Grupcev, and Yi-Cheng Tu are with the Department of Computer Science and Engineering, University of South Florida, 4202 E. Fowler Ave., ENB 118, Tampa, FL 33620, U.S.A. Emails:
akhmedovi@mail.usf.edu, ytu@cse.usf.edu}
\thanks{Joseph Fogarty and Sagar Pandit are with the Department of Physics,
University of South Florida, 4202 E. Fowler Ave., ISA 2019, Tampa,
FL 33620, U.S.A. Emails: jcfogart@mail.usf.edu, pandit@usf.edu} }

%\pubid{0000--0000/00\$00.00~\copyright~2002 IEEE}

% use only for invited papers
%\specialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}

Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on random I/O while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too.

@@

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier. In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much though it didn't bear any loss in precision and performance.

\end{abstract}

\begin{IEEEkeywords}
Push-based system, molecular simulation, scientific databases,
spatial distance histogram, GPU, parallel processing, CUDA.
\end{IEEEkeywords}
%\vspace{-5mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}\label{sc:intro}

In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

{\small
\begin{enumerate}
	\item initial physical properties are given to simulation software as arguments which are interpreted and provided in a language defined by the simulation software
	\item then it runs for given amount of time that can generally last from milliseconds to months or even more
	\item finally, the file generated during the simulation is analyzed to extract useful information
\end{enumerate}
}
\vspace{3mm}
The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software.

@@

\begin{figure}
 \centerline{ \includegraphics[width=1\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle}}
 \label{fg:collagen_dipal}
\end{figure}

\subsection{Problem Statement}
Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code.

@@

\subsection{Our approach with improvement}

@@

\subsection{Contribution and roadmap of the paper} 
The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis even faster now. Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing, which already had a good speed up in terms of performance in comparison with original pull-based approach. Since in this paper we don't expect certain form of data, we have come up with a tool that generates mock data for any number of atoms and frames, which can be used for performance tests that were done consequently, too. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU \cite{amazonGPU} nodes that consequently can be replicated and scaled up becoming more of a streaming distributed processing engine (which can be very effective on costs, since it is always easy to scale them up, shut them down and spin them back up), as well as on a simple desktop computer with a GPU device.
In general, major contribution of this paper is the framework of push-based system with an incorporated parallelism based on CUDA programming language. Besides it, there was developed a tool which generates mockup data of the same format as of the real MS simulations and represent a python script with arguments given as number of frames and atoms.

@@

\vspace{3mm}
\section{Related Work}\label{sc:relatedwork}
\vspace{3mm}

@@

\vspace{3mm}
\section{Network of Queries}\label{sc:querynetwork}
\vspace{3mm}

Having generated big amount of data in large files, now scientists need to get useful information out of it by means of analysis utilities. Often times, the queries that need to be run over the data look like selection and some kind of accumulation, if it is not more complicated. To be more exact, in this section we will try to summarize common set of queries we will be improving.

@@

\vspace{3mm}
\section{The system design}\label{sc:system}
\vspace{3mm}

The proposed design has a pretty straightforward structure which is going to be explained in this section. Shortly, the system starts with original input files (or data) and gets consumed in a sequential manner by simple disk IO framework into random access memory by chunks, then, chunk gets uploaded to GPU device and processed. At some point, the results that need to be shown to the user are transported from global memory of GPU (where they initially get accumulated) to the main memory, and then might get written as well to an output file. General data flow map is represented in Figure  \ref{fg:data-flow}.

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/memory-organization.eps} }
 \caption{Data flow in the improved design of push-based system for MS analysis}
 \label{fg:data-flow}
\end{figure}

\subsection{Data input}
Since the size of the data in the problem is huge, organization of the data flow is crucial. As mentioned before, original data is retrieved from simulation software output files and in our case it is Gromacs Software's trajectory files\footnote{The file types, their structure, and their decompression has been well explained in the previous paper in section "Data orgranization in main memory"}. For a simulation, there can be several files of different format and usually they are also compressed without any loss of information. Tool for decompressing those files to a flat format for each frame and atom has been developed in the previous work\cite{mainPaper}.

In order to experiment with different input cases, the input files have been generated with a tool that we have built in Python language. Thus, we are able to generate mock data files with any number of frames and atoms, and the values are random and can be adjusted with needs of the experiment (for example, for cases when we want to analyze bucket widths in a limited space, the coordinates are generated in range of $0 - 1$, etc.). As it has already been mentioned earlier, we don't expect the data to be distributed with a certain behavior, the importance of having this kind of tool to generate mock data is having a variety of random data sets that could reveal any issues related to values provided to the program, on top of that, having the actual file written and then being read has been a part of the experiments to have more realistic use case. In the future, the same script for generating the mock data could be used for regression and performance tests in a distributed implementation of this design as a single source.

\subsection{Data flow architecture}

MS data roughly can be represented as a list of atoms with physical properties aggregated as frames. Thus, we consider a single frame as a chunk which will be loaded into the memory\footnote{Memory in this design might be interpreted differently, since we have a separate device memory on GPUs, but it will be explained better later} and pushed through our network of query. In this case, it is very convenient, because there are almost no dependencies between frames, rather than list of atoms in the same frame, which potentially again could create additional disk I/O overhead.

Having defined the chunk, we load it from flat format file to RAM. A single atom data structure has all the attributes such as charge, mass, velocity, type of atom, coordinates, and so on. In other words, we have an array of atoms, and most of the attributes (in case of data structure, it is a class field), are simply primitive floats or doubles depending on precision requirements. Next, the same chunk of data with the same structure is loaded onto GPU global memory. GPU devices, in general, have about 4-5 types of memory, and global memory is the biggest and slowest. Consequently, depending on algorithm or query we load parts of data chunk to another types of memory called shared memory. It depends on the actual query, because the same part of chunk can be loaded and unloaded in regards with shared memory several times, especially with SDH.

\subsection{Network of queries organization}

Since we assume that our framework should be easily used by a scientist doing MS analysis, it is important to keep queries module as a separate piece of code to make it easily extendable. In order to understand how queries module is organized and the reason for it, we need to understand how CUDA and GPU work.

\emph{\textbf{GPU execution}}. From hardware standpoint, it is important to understand that every GPU device consists of SM (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such a function called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

\emph{\textbf{One-body queries kernel}}.
Commonly used CUDA array reduction\cite{gpureduction} makes the implementation of all the one-body queries obvious. Figure \ref{fg:reduction} represents general idea of the data mapping, where each cell of the vector can be associated with a thread of execution in the kernel. Having an array of atom data structures, we reduce the array accumulating all the needed data from each atom's attributes. We don't expect much performance improvement from coalesced memory access, since the data structure for atom can eliminate any gain of coalesced memory access, though for specific types of queries it could be possible to aggregate certain attributes into merged arrays and take an advantage of it. In this case, the solution should be simple in order to keep it easily extendable for other simple one-body queries for a scientist to program. We fetch all the atoms from GPU global memory, since we use references to atoms once. In this case, allocating shared memory and copying it over would complicate the implementation and possibly create additional overhead for one time reference usage. Finally, memory transportation and data consumption are implemented for user, thus, he only needs to extend the queries module which consequently can be turned into plugins model.

\emph{\textbf{Two-body queries kernel}}.
SDH (or RDF) is the algorithm that requires special treatment in our design. The point, as mentioned earlier, is that this function requires more than one iteration of the given set of atoms, but rather all combinations of pairs in a given frame, thus, it is not linear complexity. An implementation of this type of query might be very non-intuitive because of the nature of GPUs. Another issue is that since to take an advantage on this type of computation, before writing an implementation for it, a user must have a good understanding of GPU architecture which makes the system less user-friendly. At this point, it is more important to achieve good performance improvement results and then take care of ease of system extensions. As we already mentioned, GPU threads execute in parallel and consume certain amount of work followed by best try of all hardware capacities utilization, but, unfortunately, it is also important to take care of equal workload for every thread independently, so, that utilization is kept high. For this reason, we've come up with a solution that should increase performance of the computation which will be described later.

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/reduction} }
 \caption{Array reduction example\cite{gpureduction}}
 \label{fg:reduction}
\end{figure}

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/overallarchitecture} }
 \caption{Overall system architecture}
 \label{fg:overallarchitecture}
\end{figure}

\emph{\textbf{Architecture overview.}}
This section is to emphasize general architecture of the system and summarize the work flow as well as the data flow. In total, we have flat files with ready flat values for each atom aggregated in frames coming into the program as data input using common disk I/O framework of an operating system. Then, once we have loaded a frame (which is considered to be a chunk of the entire data set) onto RAM of our machine, we copy over the entire frame onto the GPU device global memory. It is important that, at this point, we never write any values to simulation files or any other data related to simulation, thus, we don't have to care about any synchronization of incoming data. After this, we have two predefined kernel functions (which were explained before), where first kernel stands for one-body queries, and the second one is for two-body query SDH. Just for architecture clearance and possible future improvements, two kernels are executed in two different streams\footnote{In CUDA stream is another abstract layer of organizing execution threads. They are not guaranteed to be executed at the same time or after each other, GPU scheduling architecture is completely different from CPU having a warp to be single scheduling unit at maximum of 32 threads at a time} (You can see them on Figure\ref{fg:overallarchitecture}). In each kernel we apply certain heuristics to improve memory access using such architecture features of GPU as shared memory and blocking of kernel threads.

\emph{\textbf{GPU grid and block sizes.}} As it has already been mentioned, utilization of hardware on GPU device is important, because it is the key to performance improvement, and the higher utilization is, the bigger improvement is.
The approach we came up with is 1 execution thread per atom. For CUDA version 2.0 maximum number of threads per block is 1024. Blocks are 3 dimensional entities in a 3 dimensional grid. The less abstract unit is, the less overhead we have in communication between them, thus, we try to utilize smallest units more. For example, for 1000,000 atoms, we would allocate 1024 threads per block as a static hard coded value, since we usually have always more than 1024 atoms in simulations. One dimension of grid can take up to 64,000 blocks, thus, if N is bigger than $1024 * 64,000$, than the rest would be allocated to $grid.y$ and $grid.z$ respectively. We have not considered numbers bigger than GPU can allocate, which could be explored in future work. It is important to understand that even though there are abstract entities of grids and blocks, it is not real hardware capacities, but they can affect scheduling and performance consequently. Decision to go with atom to thread execution is because it is pretty convenient to keep high utilization of hardware and still not lose idea if we are about to incorporate more functions, which makes the system more user-friendly, though later on as a plugin we left some room to add data transformer between RAM and GPU global memory, because there are a lot of custom ways to aggregate data and take an advantage of coalesced memory access.

\begin{figure}
 \centerline{ \includegraphics[width=1.0\columnwidth]{images/threadload} }
 \caption{SDH thread loading}
 \label{fg:threadload}
\end{figure}

\subsection{Computation work flow}

Once the data is loaded into global memory of GPU, we have two kernels for two types of queries setup for two different streams. Besides it, we allocate some memory in GPU global memory for results, too, and keep it as an object with fields to easily pull it back when it is ready, in our case all the data sizes are static before the kernels are executed, which follows nature of GPU devices. 
\subsubsection{One-body kernel} Every thread of execution in one-body query writes into shared memory, before accumulating the data into global memory result in order to avoid massive read and write blocking conflicts which will eliminate any kind of performance advantages. Once a block is finished, first thread of each block will accumulate data from shared memory into global synchronizing the results, every write operation to shared and global memories is atomic (e.g. generic CUDA function atomicAdd). We run the first stream with one-body queries and wait till it gets finished. The reason is that we expect certain behavior from shared memory from each of the streams, thus, just to make sure that they don't overlap, we run them separately. While the computation is running, there is a possibility to use a third stream to load another chunk of data, too. First stream and kernel execute simple vector reduction type of computation accumulating all the needed data and keep it in the GPU.
\subsubsection{Two-body kernel} Once the first kernel is finished we run the two-body kernel with its own stream. As it has already been mentioned, the biggest workload is in this kernel, because it has higher complexity. In this particular case it is $O(n^2)$. GPU streaming multiprocessors are very powerful for straightforward primitive independent calculations, and each core is much weaker than a moden CPU which has a whole bunch of L caches with some prediction behavior systems for code divergence. As you can see on Figure \ref{fg:threadload}, unbalanced thread load assumes for GPU kernel to check every time if the target atom can interact with other atom and if it has to wait till other threads finish their jobs. Thus, we had to come up with a strategy of making equally balanced thread jobs and distribute them among the threads leaving atom to thread mapping as is.

For SDH we the following strategy

\begin{itemize}
\item[1.] Find current index (for target atom)
\item[2.] Find range of atoms it has to interact with in an equally loaded fashion
\item[3.] Load the range of atoms into shared memory
\item[4.] Iterate through the range of atoms from shared memory and atomically write the distances counts into buckets pre-allocated in share memory
\item[5.] Synchronize with other threads in the block (not in the entire grid)
\item[6.] If it is the first thread in block, add all bucket counts from shared memory into global memory
\end{itemize}

\section{Experiments Results}\label{sc:experiments}

The framework has been developed in a programming language CUDA, and the sequential version for benchmark\cite{mainPaper} has been originally developed C++. Experiments were run on operating systems which didn't have any additional workload besides usual installation, and the operating systems particularly were Amazon Linux, which doesn't have a lot of modifications except for some additional features for management with Amazon EC-2 Console. We have generated mock data sets with different variations of atoms, frames and ran them with different workloads comparing sequential version of Push-based system with improved version written in CUDA.

\emph{\textbf{Hardware stack.}}
There are primarily several parts that are really important in this hardware stack. Since this is primarily big data processing, we care about processing power and memory. RAM on the node that we were running has 16GB. The server allocated from Amazon has the following CPU parameters:

\begin{center}
  \begin{tabular}{ l  l }

Architecture & x86\_64 \\  
CPU op-mode(s) & 32-bit, 64-bit \\  
Byte Order & Little Endian \\  
CPU(s) & 8 \\  
On-line CPU(s) list & 0-7 \\  
Thread(s) per core & 2 \\  
Core(s) per socket & 4 \\  
CPU family & 6 \\  
Model & 45 \\  
Model name & Intel Xeon E5-2670 2.60GHz \\  
Stepping & 7 \\  
CPU MHz & 2593.814 \\  
L1d cache & 32K \\  
L1i cache & 32K \\  
L2 cache & 256K \\  
L3 cache & 20480K \\  
NUMA node0 CPU(s) & 0-7 \\

  \end{tabular}
\end{center}

The same CPU has been used for sequential as well as for improved GPU version during experiments with benchmarks. Just to mentioned, CPU in improved version is used in-between memory loading from file and uploading it to GPU, printing the results to the user, which is not big load. Consequently, we also have GPU device with the following parameters:

\begin{center}
  \begin{tabular}{ l l }
Name & Grid 34C K520 \\
Memory & 4095Mib \\ 
Total Nvidia CUDA Cores & 1536 \\ 
Memory Size & 8 GB GDDR5 (4 GB/GPU) \\
Total GPUs & 2 GK104 GPUs \\
  \end{tabular}
\end{center}

\emph{\textbf{Data sets.}} As it has already been mentioned, we have developed a small python script that generated data sets that are used for the experiments. It accepts two numbers as arguments (number of atoms and frames respectively) and generate a file with appropriate number of atoms/frames and format. We have generated number of mock data sets in range of $1K-8M$ atoms with $10, 100, 1000$ frames each. For each data set we run workloads with the values: $1, 5, 25, 50, 100$ . Number of atoms and frames are obviously up to the scientist and depend on simulations, but workloads are a bit trickier. First of all, they give us a bit better understanding of the average value, which might vary on different circumstances related to OS, secondly, once this system is integrated into a distributed system, we can imagine that each frame could be processed by a separate node keeping the values inside of the appropriate memory (global or RAM) and generate responses based on the clients selections requests. This possibility will be elaborated in more details in Future Work section.

\emph{\textbf{SDH implementation.}} Even though implementation of SDH can be different, in the previous paper the algorithm in the previous paper was based on quad tree data structure with data proximity approach based on data uniformity\cite{mainPaper}, but in our experiments we follow special CUDA based approach which perfectly fits the nature of GPU devices with the given memory hierarchy, and the number of computations are equal to brute force, consequently, there is no expectation on data uniformity, thus, the problem solution complexity is still the same with the worst case of $O(N^2)$, but processed in parallel by GPU cores. More improvements can be done for acceleration of this approach, but in the system we have not gone too far with data converter module, because initial framework has to be simple and user friendly.

\subsection{Results}
Results are represented, in general, on Figures \ref{fg:chart_1000}, \ref{fg:chart_10000} \ref{fg:chart_max}. You can instantly notice that speedup for 1000 atoms is not very impressive, especially for $10$ frames, and the reason for that is nature of GPU and memory transfer. There is a small overhead we have to consider before computing part: in order to let GPU process the data, we have to transfer the data to Global Memory, and upon completion it is needed to load the results back to RAM to show it to the user. In this particular case, for small workload and frames, it is even slower than the sequential processing. Afterwards, the bigger workload the more convincing speedup gets, unfortunately, it doesn't get more than $4$ times for 1000 atoms. Fortunately, scientists usually work with much larger data sets, thus, this doesn't really represent general speedup for average use case.

\emph{\textbf{10000 atoms.}}
As you can see, speedup bumps up to almost 30 with 10000 atoms (Figure \ref{fg:chart_10000}). Once the overhead on memory transfer is covered by computation speedup, it keeps growing. It is important to pay attention to the growth of speedup over growth of workload. It is not that significant, because even though CPUs are much slower, they still use smart caching and internal multi-core processing, which is pretty fast, too.

\begin{figure}
 \centerline{ \includegraphics[width=1.0\columnwidth]{images/chart_1000}}
 \caption{Speedup for 1000 atoms}
 \label{fg:chart_1000}
\end{figure}

\begin{figure}
 \centerline{\includegraphics[width=1.0\columnwidth]{images/chart_10000}}
 \caption{Speedup for 10000 atoms}
 \label{fg:chart_10000}
\end{figure}

Fortunately, scientific data is much bigger than above examples. You can see average speedup for bigger number of atoms on Figure \ref{fg:chart_max}. The speedup gets upto 70 times which could save huge amount of time for scientists. These results are not the best, as it has been already mentioned earlier, since we have not modified general structure of memory being transferred to Global Memory not taking advantage of coalesced memory access to make it easily extendable.

\begin{figure}
 \centerline{\includegraphics[width=1.0\columnwidth]{images/chart_max}}
 \caption{Speedup for larger number of atoms}
 \label{fg:chart_max}
\end{figure}

It is noticable that growth of the speedup is not stable and with certain number of atoms it drops dramatically. It is because we've used static approach of computing grid dimensions. Besides it, in the future, it can be expanded into a dynamic version that will automatically adjust the grid dimensions choosing least overhead that it has to go with.

\section{Conclusions and Future Work}\label{sc:conclusion}
The ultimate goal of this work was to get an improved version of Push-based system for molecular simulation data analysis. As you could see from result on charts (Figure \ref{fg:chart_1000}, \ref{fg:chart_max}), the proposed improved version of the system could save a lot of time for scientists which is up to 60-70 times less of time. The results have been gathered with a single snapshot of the framework tracked in control version system which can be improved with means of data converter module and dynamic grid dimensions computation module. In the future, this system has all potentials to be expanded into a distributed system, where frames could be aggregated by nodes and processed in parallel.\\

\bibliographystyle{IEEEtran}
\bibliography{references.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendices

>>>>

\documentclass[11pt,a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[dvips]{graphicx}
\usepackage{xcolor}
\usepackage{times}
\usepackage{ragged2e}
\usepackage{titlesec}
\usepackage{wrapfig,subfigure,setspace,verbatim,cite,url,graphicx,float}
\usepackage{algorithmic,amsfonts,amssymb,amsmath,calc,calrsfs,cleveref}
\usepackage[justification=centering]{caption}

\usepackage[
breaklinks=true,colorlinks=true,
%linkcolor=blue,urlcolor=blue,citecolor=blue,% PDF VIEW
linkcolor=black,urlcolor=black,citecolor=black,% PRINT
bookmarks=true,bookmarksopenlevel=2]{hyperref}

\usepackage{geometry}
% PDF VIEW
\geometry{total={210mm,297mm},
left=25mm,right=25mm,%
bindingoffset=0mm, top=25mm,bottom=25mm}
%PRINT
\geometry{total={210mm,297mm},
left=20mm,right=20mm,
bindingoffset=0mm, top=15mm,bottom=35mm}

\linespread{1.5}

\titleformat{\chapter}
{\normalfont\LARGE\bfseries}{\thechapter.}{1em}{}

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%

@@

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
%   TITLEPAGE
%
%   due to variety of titlepage schemes it is probably better to make titlepage manually
%
%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
\thispagestyle{empty}

{%%%
\sffamily
\centering

{\Large 
Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU
}

\vspace{1.5cm}

{
by
}
\vspace{.5cm}

{ 
Iliiazbek Akhmedov
}

\vspace{3.5cm}

A thesis submitted in partial fulfillment

of the requirements for the degree of

Master of Science in Computer Science

Department of Computer Science and Engineering

College of Engineering

University of South Florida

\vspace{3.5cm}
Major Professor:  Yicheng Tu, Ph.D.
\vspace{\fill}

\vspace{.5cm}
Date of Approval:

N/A
\vspace{\fill}

Copyright \textcopyright\space 2016, Iliiazbek Akhmedov

%%%
}%%%

\pagestyle{plain}
\centering

\textbf{DEDICATION}
\vspace{0.5cm}

\begin{flushleft}
\hspace{3em} I dedicate my thesis work to my family who have always been supporting me throughout my life. I also dedicate it to my closest friends who have shared my passion and encouraged me to explore and try my ideas in life.
\end{flushleft}

\pagestyle{plain}
\centering
\textbf{ACKNOWLEDGMENTS}
\vspace{0.5cm}

\begin{flushleft}
\hspace{3em} I would like to thank my major professor throughout my master thesis Dr. Yicheng Tu for his guidance and knowledge that I received through all the time I worked with him. I really appreciate his contribution and very happy that I got a chance to work with him. I as well want to thank the committee members for their support and encouragement.
\end{flushleft}

\tableofcontents

\begin{abstract}
\justify
Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on random I/O while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too.

@@

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier. In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much though it didn't bear any loss in precision and performance.
\end{abstract}

%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%
%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%---%%%

\chapter{Introduction}

\justify
In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

\begin{enumerate}
	\item initial physical properties are given to simulation software as arguments which are interpreted and provided in a language defined by the simulation software
	\item then it runs for given amount of time that can generally last from milliseconds to months or even more
	\item finally, the file generated during the simulation is analyzed to extract useful information
\end{enumerate}

\vspace{3mm}
The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software.

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle}}
 \label{fg:collagen_dipal}
\end{figure}

\section{Problem Statement}
Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code.

@@

\section{Our approach with improvement}

@@

\section{Contribution and roadmap of the paper} 
The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis even faster now. Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing, which already had a good speed up in terms of performance in comparison with original pull-based approach. Since in this paper we don't expect certain form of data, we have come up with a tool that generates mock data for any number of atoms and frames, which can be used for performance tests that were done consequently, too. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU \cite{amazonGPU} nodes that consequently can be replicated and scaled up becoming more of a streaming distributed processing engine (which can be very effective on costs, since it is always easy to scale them up, shut them down and spin them back up), as well as on a simple desktop computer with a GPU device.
In general, major contribution of this paper is the framework of push-based system with an incorporated parallelism based on CUDA programming language. Besides it, there was developed a tool which generates mockup data of the same format as of the real MS simulations and represent a python script with arguments given as number of frames and atoms.

@@

\chapter{Related Work}\label{sc:relatedwork}

@@

\section{Network of Queries}\label{sc:querynetwork}
Having generated big amount of data in large files, now scientists need to get useful information out of it by means of analysis utilities. Often times, the queries that need to be run over the data look like selection and some kind of accumulation, if it is not more complicated. To be more exact, in this section we will try to summarize common set of queries we will be improving.

@@

\bibliographystyle{unsrt}
\bibliography{references.bib}
